---
title: "R Data manipulation and visualization"
subtitle: "EXGEN5449 - Winter 2019"
author: "Stefan Schreiber"
output:
  html_document:
    theme: paper
    highlight: pygments
    toc: true
    number_sections: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
---

<script>
$(document).ready(function() {
  $items = $('div#TOC li');
  $items.each(function(idx) {
    num_ul = $(this).parentsUntil('#TOC').length;
    $(this).css({'text-indent': num_ul * 10, 'padding-left': 0});
  });

});
</script>

```{css echo=FALSE}
.red { 
  color: red;
  padding-top: 0;
  margin-top: -20px;
  border-top-color: #f5f5f5;
}
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

# Tidyverse
Today we are going to start using the [tidyverse](https://www.tidyverse.org/) packages for data manipulation and visualization. According to the tidyverse developers:

> The tidyverse is an opinionated collection of R packages designed for data science. All packages share an underlying design philosophy, grammar, and data structures. 


To install the tidyverse packages you can simply type `install.packages("tidyverse")` or you can  install it by clicking the `Packages` tab in R Studio (right hand side), then click "Install", find the tidyverse package and start the installation.

Before we continue let's make sure to load the tidyverse package:

```{r message=FALSE}
require(tidyverse) #OR
library(tidyverse)
```

Alright with that being out of the way let's head over to the [tidyverse](https://www.tidyverse.org/) website and start exploring what we can do! Then start working on a few examples from the [R for Data Science](https://r4ds.had.co.nz/) book.

# Maps for Tomas
## Using latitude and longitude data
```{r message=FALSE}
# load packages
require(maps)
require(mapdata)

# convert map data format to data frame using `map_data()` function
w2hr <- map_data("world2Hires")
head(w2hr)

# conver to tibble (not necessary but follows the tidyverse philosophy)
w2hr <- as_tibble(w2hr)

# subset Canada
w2hr %>% filter(region == "Canada") -> w2hr_can

# plot
ggplot() +
  geom_polygon(data = w2hr_can, aes(x = long, y = lat, group = group),
               fill = "white", colour = "black") +
  coord_quickmap()

# add new data sets with lats and longs which can be used to map on top
can_cities <- tibble(lat = c(56.732,53.631,51.044,49.699),
                     long = c(-111.376,-113.324, -114.062, -112.818),
                     city = c("Fort McMurray", "Edmonton", "Calgary", "Lethbridge"))

# change longitudes from the `can_cities` tibble (-180 to +180),
# so that it matches the 0-360 degree format in the `w2hr` map dataset
can_cities %>% 
  mutate(long = long + 360) -> can_cities2

# plot
ggplot() +
  geom_polygon(data = w2hr_can, aes(x = long, y = lat, group = group),
               fill = "white", colour = "black") +
  coord_quickmap()+ 
  geom_point(data = can_cities2, aes(x = long, y = lat)) +
  geom_text(data = can_cities2, aes(x = long + 5, y = lat + .8, label = city))

```

**Excercise**: Can you find the population size for each cities and add this to the tibble `can_cities`? Then use this information to let the size of the poplulation be shown as differently sized points.


## Using Google maps
There is a great R package called `ggmap` which allows you to make fantastic maps in R. Please have a look at the [developers source site](https://github.com/dkahle/ggmap) as it has important information. Furthermore have look at [this article](https://journal.r-project.org/archive/2013-1/kahle-wickham.pdf) in the R Journal about the `ggmap` package. It is a little "dated" (2013) but will still give you an idea what is possible.
